Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-07-05 Manuscript.qmd
1st submission

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_2024OSF]). The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.

@@

Another shell whose results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io]. In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub). These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August). Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88 (@fig-ORK_sub).

@@

These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past. Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl]. By using LIBS as a screening tool for targeted \delta^{18}O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency. This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.

>>>>

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_Surge_Godino_2024]). The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.

@@

Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io]. In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub). These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August). Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88 (@fig-ORK_sub).

@@

These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past. Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl]. By using LIBS as a screening tool for targeted $\delta$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency. This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.
